Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-07-11 AnalyticStructure_main.tex
Update on Overleaf.

% special operators
\newcommand{\dmnot}{\op{\rho}_0}
\newcommand{\dm}{\op{\rho}}
\newcommand{\hnot}{\op{H}_0}
\newcommand{\hone}[1]{\op{H}_1\argument{#1}}
\newcommand{\transition}[1]{\op T_{#1}}
\newcommand{\excite}[2]{\op e_{{#1}{#2}}}
\newcommand{\decay}[2]{\op d_{{#1}{#2}}}
\newcommand{\Liouv}{\sop{\mathcal L}}
\newcommand{\Liouvnot}{\sop{\mathcal L_0}}
\newcommand{\coupl}{\sop{\mathcal K}}
\newcommand{\honetmp}[1][]{\op{H_1}\argument{#1}} 
\newcommand{\identity}{\op{\mathbb I}}
\newcommand{\rmat}[1]{\optr R}

@@

%\title{On the analytic structure of the linear response susceptibility in open systems}
\title{Localization properties, observable character and computational reliability of the linear response in open systems}

\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), SP2M, INAC, CEA-UJF, 17 Av. des Martyrs,
38054 Grenoble, France}
\author{Marco D'Alessandro}%
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}

\date{\today}

\begin{abstract}
We analyze how the fundamental equations which govern the linear response properties of a quantum system are influenced by the presence of open boundaries.
In particular we focus our attention on linear-response (LR)-TDDFT calculation of molecular systems. We show that optical excitations of energy below the LR ionization potential 
behave as observable, localized quantities, that can be computationally calculated with high precision, provided an adequate level of completeness.
Above this energy threshold, the analytic structure of the susceptibility functional might be separated in two sectors: one can still be described in terms of observable excitation 
with discrete energies, that can be captured within computational setups tailored for localized states. However such excitations are embedded in a continuum of excited states 
that, on the contrary, behave as plane waves and cannot be described efficiently by the same computational techniques. Such excitations do not exhibit observable features and 
\emph{cannot} be considered as poles of the dynamical polarizability. Nonetheless, the contribution of such continuum of excitation reveals to be of paramount importance even 
for expressing LR quantities in the infrared regime like for example static polarizability.
This is particularly evident for small molecules; we show how these considerations might be influenced by the size of molecular system.
We then present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme.  This result is a inherent behavior of the system's Liouvillian, and does not depend on computational treatment of the 
unoccupied subspace.
\end{abstract}

@@

A recent, remarkable example of precision-driven initiative is provided by the DeltaCodes project~\cite{deltaTest2016},
where a grassroot community of code-owners struggled to extract well-defined computational results in the context of 
Density Functional Theory (DFT) calculations. Ground-State quantum mechanical quantities as lattice constants and bulk modulus of 
elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties 
on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code. 
Any other code able to extract the same quantities with the same approximations might be compared with the 
results of the community in order to assess its computational reliability.

@@

On the other hand, positive energy solutions behave differently. 
Even though the potential is localized, their positive energy makes them behave 
as spherical waves and its convolution with the source term  gives raise to a delocalized wavefunction $\psi(\r)$ 
in the whole accessible space.
As they cannot be normalized they belong to the essential spectrum of the hamiltonian operator and form a \emph{continuum} of states.
Thanks to the Lippman-Schwinger relation of Eq.~\eqref{SEHelmholtz1} they can be put in bijection with the eigenstates of the 
free-particle hamiltonian $\op H_s$. We may say that their quantum numbers are determined by the Laplacian operator, 
regardless of the particular features of the potential.

@@

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}. The authors of this paper analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, 
and the independence of its energy with respect to the basis set. It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
Furthermore, the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.

@@

The response density operator is expressed as the first-order variation of $\dmnot=\sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$, where the set $\{\ket{\psi_p}\}$ denote the 
occupied states, \emph{i.e} the subset of the bound eigenstates\footnote{In what follows we will assume that the set of states $\{\ket{\psi_p}\}$ are described by real 
functions, when projected in the $\r$-representation.} of $\hnot$ with (negative) energy $\eps_p$ lower than the Fermi level. Using this notation we denote the energy of the HOMO level 
as $\eps_h$ and identify the value of the first ionization potential with $|\eps_h|$.

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm'_\Phi = \dm_{\perp} \doteq \dmnot\dm'_\Phi\op Q_0 + \op Q_0\dm'_\Phi\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} 
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with%:here in this case the source term contains further addend apart from the unperturbed potential and reads:
$%\be
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p}
$. %\ee

It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that govern the localization properties of the corresponding fluctuation state.
As a \emph{stringent} hypothesis, let us now assume that the ket $\ket{s_p^\Phi(\omega)}$ is localized, namely its evaluation can be
restricted (if suitable) on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
%in the $\r$-representation. 
Even when such locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of 
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range, 
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s(\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the 
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian can be extracted with traditional GS techniques, 
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability (blue continuous line) and of the ground state permanent dipole (red dotted line) 
of the $CO$ molecule with respect to the size of the computational domain. The inset plot evidences the relative percentage convergence using the value associated to the largest 
box as reference.}
\end{figure}

From the linearized Schr\"odinger Equation of the perturbed problem, it is easy to see that the fluctuation state at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behavior of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than $|\eps_p|$, this is consistent with the above considerations.

% A direct confirmation of the localization character of fluctuation states at $\omega=0$ can be achieved through a ground-state calculation. Indeed, $\ket{f_p(\omega)}$ are nothing but the 
% perturbed KS eigenstates projected out from the domain of $\dmnot$, so for a constant perturbation $\op\Phi_s$ it turns out that $\ket{f_p(0)} = \op Q_0\ket{\psi_p^{\Phi_s}}$. The perturbed KS 
% states $\ket{\psi_p^{\Phi_s}}$ can be computed (beyond LR) by performing a DFT ground-state computation in which the constant perturbing field is added to the external potential and the 
% spatial extension of the fluctuation states can be directly measured. 
In Fig.~\eqref{co_alphaStatic} illustrate this procedure for a $CO$ molecule 
perturbed by a static electric field. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The plots evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals.

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions. In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
\emph{i.e.} inside the scalar products of 
\eqref{LinearResponseFunctDef2}. 
In this case $\ket{f_p^\Phi(\omega)}$ is a unbound, delocalized function; a numerical treatment tailored for localized
functions will \emph{never} be able to discretise such function with arbitrary accuracy.  
In this context, the bound state character of $\{\bra{\psi_p}\op O\}$ behave as a regulator:
\emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
This observation indicates that, above threshold, a computational setup 
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able (even implicitly) to express an
unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.
The non-vanishing spherical-wave behavior in the long range of $\ket{f_p^\Phi(\omega)}$ is therefore responsible
of this convergence difficulties, especially for high frequencies.

As an explicit example in support of these arguments we compare the absorption spectrum of a $CO$ molecule computed in 
three distinct real-space computational setups, which differs for the dimension of 
the computational domain. Results reported in figure \ref{co_spectrum} evidences that the curves are almost coincident in the below threshold regime,
whereas the effect of the computational setup are clearly visible for values of $\omega$ that exceed the threshold value.
For these ultraviolet regions, convergence might be reached for domains of much larger sizes than what presented here~\cite{baroni2008},
which in turn reveal more than sufficient below threshold.

@@

In view of reproducibility, the scalar product of Eq.~\eqref{LinearResponseFunctDef1}, defining the observable fluctuation, may be compared among various numerical treatments.
However, even though directly related to experimental results, such quantity is not by itself a property of the molecule, as it is a function of $\omega$ that explicitly depends 
on the perturbation operator $\op \Phi$, not to mention the target observable $\op O$. Alternatively, we may consider only the fluctuation state, but it is still a $\op \Phi$-dependent wavefunction, 
%that may be local and with a well-defined normalization 
with a potential local behavior only for $\omega$ below the ionization threshold. So, Comparing fluctuation states among diverse numerical treatments would be a 
challenging task.

@@

In Fig.~\ref{CO_exc} and \ref{C6H6_exc} we plot the Density of States of the Liouvillian for the $CO$  and Benzene molecules, 
for two different numerical setups of increasing simulation domains. 
The figures show clearly that all the excitations of energy $\Omega_a$ lying below the 
molecule ionization potential (IP) have a clear localized behaviour, as their energy does not vary with the size of the simulation domain.
On the other hand, above the first ionization threshold the excitation energies have an energy that strongly depends on the computational
setup, which is in line with the expected pseudo-continuum character of this sector. 
In addition, for the benzene molecule it is also possible to 
identify excitations of energy above IP that still possess a localized behaviour: these excitations satisfy the
condition $\Omega_a < |\eps_a|$ presented in the above section.
In Fig.~\ref{c6h6_excLand} we collect the excitations of the benzene molecule following this criterion:
the agreement with the expected analytic structure presented in Fig.~\ref{ExcitationLandscape} is impressive.

@@

The situation is not so simple in the above-threshold regime.
In this case what is important is matrix element of the observable of interest
between the bound, occupied state and the delocalized fluctuation state.
To achieve convergence, care should be taken in expressing accurately the FS on the
appropriate region of space. In view of reproducibility, computational setups which are able to express precisely
delocalized states seem more interesting in this regime.

@@

\section{Criterion of computational reliability for the numerical convergence of the LR}

>>>>

% special operators
\newcommand{\dmnot}{\op{\rho}_0}
\newcommand{\dm}{\op{\rho}}
\newcommand{\hnot}{\op{H}_0}
\newcommand{\hone}[1]{\op{H}_1\argument{#1}}
\newcommand{\transition}[1]{\op T_{#1}}
\newcommand{\excite}[2]{\op e_{{#1}{#2}}}
\newcommand{\decay}[2]{\op d_{{#1}{#2}}}
\newcommand{\Liouv}{\sop{\mathcal L}}
\newcommand{\Liouvnot}{\sop{\mathcal L_0}}
\newcommand{\coupl}{\sop{\mathcal K}}
\newcommand{\honetmp}[1][]{\op{H_1}\argument{#1}} 
\newcommand{\identity}{\op{\mathbb I}}
\newcommand{\rmat}[1]{\optr R}
\newcommand{\fscd}[1]{\ket{f_p^\Phi(\omega)^{#1}}}

@@

%\title{On the analytic structure of the linear response susceptibility in open systems}
\title{Locality and Computational Reliability of Linear Response Calculations for Molecular Systems}% Localization properties, observable character and computational reliability of the linear response in open systems}
\author{Marco D'Alessandro}%
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}
\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), SP2M, INAC, CEA-UJF, 17 Av. des Martyrs,
38054 Grenoble, France}
\date{\today}

\begin{abstract}
By performing a critical analysis of the fundamental equations of linear-response  (LR) formalisms in molecules, 
we explore the interplay between locality of the response density operator and  numerical convergence of LR-related quantities.
We show that the first ionization potential of the system behaves as a threshold energy: for frequencies below this value, it is possible to parametrize the response density by employing localized states only. Above this threshold, such a locality property cannot be achieved. 
Such considerations may be transposed in terms of the molecule's excited states. We show that not all the system's excitations can be considered on equal footing.
There is a discrete sector of excitations that can be parametrized by observable, localized states, which can be computationally expressed with high precision, provided an adequate level of completeness. 
We present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme.  
The remaining excitation modes belong to a continuum spectrum that, on the contrary, is not directly associated to observable properties and can only be effectively represented in a given computational setup.
Such considerations are important not only for reproducibility of the results among different computer codes 
employing diverse formalisms, but also in view of providing a deeper understanding on the impact of models' 
approximations on the scientific outcomes of the simulation.

% ---
% We discuss these aspects in the context of linear response treatment of molecular systems. 
% By analysing simple case-studies with a highly complete numerical formalism, 
% we explore the interplay between locality of the perturbed wavefunctions and 
% numerical convergence of observable quantities. This will enable us to draw some considerations on the analytic structure of the linear susceptivity for open systems and on the potential convergence issues that might arise in a given computational setup.

% We believe that such discussion could provide useful considerations and perspectives for the design of a effective approaches for the study of molecular excitations in realistic conditions.

% We analyze how the fundamental equations which govern the linear response properties of a quantum system are influenced by the presence of open boundaries.
% In particular we focus our attention on linear-response (LR)-TDDFT calculation of molecular systems.
% Above this energy threshold, the excitation spectrum may be separated in two sectors: one can still be described in terms of observable excitation 
% with discrete energies, that can be captured within computational setups tailored for localized states. 
% This result is a inherent behavior of the system's Liouvillian, and does not depend on computational treatment of the 
% unoccupied subspace.
\end{abstract}

@@

A recent, remarkable example of precision-driven initiative is provided by the DeltaCodes project~\cite{deltaTest2016},
where a grassroot community of code-owners struggled to extract well-defined computational results in the context of 
Density Functional Theory (DFT) calculations. Ground-State quantum mechanical quantities as lattice constants and bulk modulus of 
elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties 
on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code. 
Any other code able to extract the same quantities with the same approximations may be compared with the 
results of the community in order to assess its computational reliability.

@@

On the other hand, positive energy solutions behave differently. 
Even though the potential is localized, their positive energy makes them behave 
as spherical waves and its convolution with the source term  gives raise to a delocalized wavefunction $\psi(\r)$ 
in the whole accessible space.
As they cannot be normalized they belong to the essential spectrum of the Hamiltonian operator and form a \emph{continuum} of states.
Thanks to the Lippmann-Schwinger relation of Eq.~\eqref{SEHelmholtz1} they can be put in bijection with the eigenstates of the 
free-particle Hamiltonian $\op H_s$. We may say that their quantum numbers are determined by the Laplacian operator, 
regardless of the particular features of the potential.

@@

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}. The authors of this paper analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, 
and the independence of its energy with respect to the basis set. 
% It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
% able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
% true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
On the contrary, it is clearly shown that the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.

@@

The response density operator is expressed as the first-order variation of $\dmnot=\sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$, where the set $\{\ket{\psi_p}\}$ denote the 
occupied states, \emph{i.e} the subset of the bound eigenstates\footnote{In what follows we will assume that the set of states $\{\ket{\psi_p}\}$ are described by real 
functions, when projected in the $\r$-representation.} of $\hnot$ with (negative) energy $\eps_p$ lower than the Fermi level. Using this notation we denote the energy of the HOMO level 
as $\eps_h$ and identify the value of the first ionization potential with IP~$\equiv|\eps_h|$.

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part plus a coupling term. The action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$.

The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm'_\Phi = \dm_{\perp} \doteq \dmnot\dm'_\Phi\op Q_0 + \op Q_0\dm'_\Phi\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} 
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with%:here in this case the source term contains further addend apart from the unperturbed potential and reads:
$%\be
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p}
$. %\ee

It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that govern the localization properties of the corresponding fluctuation state.
As a \emph{stringent} hypothesis, let us now assume that the ket $\ket{s_p^\Phi(\omega)}$ is localized, namely its evaluation can be
restricted on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
%in the $\r$-representation. 
Even when such locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of 
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range, 
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the 
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian can be extracted with traditional GS techniques, 
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability (blue continuous line) and of the ground state permanent dipole (red dotted line) 
of the $CO$ molecule with respect to the size of the computational domain. The inset plot evidences the relative percentage convergence using the value associated to the largest 
box as reference.}
\end{figure}

From the linearized Schr\"odinger Equation of the perturbed problem, it is easy to see that the fluctuation state at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behavior of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than any of the $|\eps_p|$, this is consistent with the above considerations.

Such observations enable us to express quantities like the static polarizability tensor $\alpha_{ij}$ from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} = 
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
% A direct confirmation of the localization character of fluctuation states at $\omega=0$ can be achieved through a ground-state calculation. Indeed, $\ket{f_p(\omega)}$ are nothing but the 
% perturbed KS eigenstates projected out from the domain of $\dmnot$, so for a constant perturbation $\op\Phi_s$ it turns out that $\ket{f_p(0)} = \op Q_0\ket{\psi_p^{\Phi_s}}$. The perturbed KS 
% states $\ket{\psi_p^{\Phi_s}}$ can be computed (beyond LR) by performing a DFT ground-state computation in which the constant perturbing field is added to the external potential and the 
% spatial extension of the fluctuation states can be directly measured.
where $\mathbf d^\text{(GS)}$ is the ground state dipole.
In Fig.~\eqref{co_alphaStatic} illustrate this procedure for a $CO$ molecule 
perturbed by a static electric field. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The plots evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals.

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions. In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
\emph{i.e.} inside the scalar products of 
\eqref{LinearResponseFunctDef2}. 
Since $\ket{f_p^\Phi(\omega)}$ becomes an unbound, delocalized function, the bound state character of $\{\bra{\psi_p}\op O\}$ behave as a regulator:
\emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
This observation indicates that, above threshold, a computational setup 
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able (even implicitly) to express an
unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.
The non-vanishing spherical-wave behavior in the long range of $\ket{f_p^\Phi(\omega)}$ is therefore responsible
of this convergence difficulties, especially for high frequencies.

As an explicit example in support of these arguments we compare the absorption spectrum of a $CO$ molecule computed in 
three distinct real-space computational setups, which differs for the dimension of 
the computational domain. Results reported in figure \ref{co_spectrum} evidences that the curves are almost coincident in the below threshold regime,
whereas the effect of the computational setup are clearly visible for values of $\omega$ that exceed the threshold value.
For these high-frequency regions, convergence might be reached for domains of much larger sizes than what presented here~\cite{baroni2008},
which in turn reveal more than sufficient below threshold.

@@

In view of reproducibility, the scalar product of Eq.~\eqref{LinearResponseFunctDef1}, defining the observable fluctuation, may be compared among various numerical treatments.
However, even though directly related to experimental results, such quantity is not by itself a property of the molecule, as it is a function of $\omega$ that explicitly depends 
on the perturbation operator $\op \Phi$, not to mention the target observable $\op O$. Alternatively, we may consider only the fluctuation state, but it is still a $\op \Phi$-dependent wavefunction, 
%that may be local and with a well-defined normalization 
with a potential local behavior only for $\omega$ below the ionization threshold. Hence, comparing fluctuation states among diverse numerical treatments would be a 
challenging task.

@@

In Fig.~\ref{CO_exc} and \ref{C6H6_exc} we plot the Density of States of the Liouvillian for the $CO$  and Benzene molecules, 
for two different numerical setups of increasing simulation domains. 
The figures show clearly that all the excitations of energy $\Omega_a$ lying below the 
molecule ionization potential (IP) have a clear localized behaviour, as their energy does not vary with the size of the simulation domain.
On the other hand, above the first ionization threshold the excitation energies have an energy that strongly depends on the computational
setup, which is in line with the expected pseudo-continuum character of this sector. 
In addition, for the benzene molecule it is also possible to 
identify excitations of energy above IP that still possess a localized behaviour: these excitations satisfy the
condition $\Omega_a < |\eps_a|$ presented in the above section.
In Fig.~\ref{c6h6_excLand} we collect the excitations of the benzene molecule following this criterion:
the agreement with the expected analytic structure presented in Fig.~\ref{ExcitationLandscape} is remarkable.

@@

The situation is not so simple in the above-threshold regime.
In this case what is important is the matrix element of the observable of interest
between the bound, occupied state and the fluctuation state (FS). 
The FS is here an unbound, delocalized function, and a numerical treatment tailored for localized functions will \emph{never} be able to discretise such function with arbitrary accuracy.  
To achieve numerical convergence, care should be taken in expressing accurately the FS on the
appropriate region of space. In view of reproducibility, computational setups which are able to express precisely
delocalized states seem more interesting in this regime.

@@
